Making Soup: Preparing and Validating Molecular Simulations of the Bacterial Cytoplasm

### 2019-09-30 paper/virtual_cytoplasm_v2.tex
final changes

\begin{abstract}
Biomolecular crowding affects the biophysical and biochemical behavior of macromolecules when compared to the dilute environment present in experiments made with isolated proteins. Computational modeling and simulation are essential tools in the investigation of how crowding affects the structural dynamics and biological activities of macromolecules. As computational power increased, modeling and simulating large scale all-atom explicit solvent models of the prokaryote cytoplasm become possible. In this work, we build an atomistic model of the cytoplasm of \textit{Escherichia coli} composed by 1.5 million atoms and submit it to a total of \SI{3}{\micro\second} of molecular dynamics simulations, which is unprecedented for a system of this scale. Our model reproduces experimental data about the diffusion coefficient and stability of macromolecules under crowded conditions. We provide a Python script and a set of files that enables any researcher to build their own \textit{E. coli} cytoplasm models to address questions related to crowding.
%For Nature, the abstract is really an introductory paragraph set
%in bold type.  This paragraph must be ``fully referenced'' and
%less than 180 words for Letters.  This is the thing that is
%supposed to be aimed at people from other disciplines and is
%arguably the most important part to getting your paper past the
%editors.  End this paragraph with a sentence like ``Here we
%show...'' or something similar.

@@

The intracellular microenvironment is markedly different from the \textit{in vitro} simplification from which most of our current knowledge about the biochemical and biophysical behavior of macromolecules was derived~\cite{Feig2017a}. Instead of simple and diluted, the \textit{in vivo} microenvironment is highly heterogeneous and concentrated, leading to spatial constrains that directly affects biological activities~\citep{ostrowska2019}. This effect, called macromolecular crowding, directly influences the stability of macromolecules through an equilibrium between stabilization of their native structure due to the entropic component given by the excluded volume effect~\cite{cheung2005} and destabilizationg due to enthalpic component given by unspecific interactions with the crowders~\cite{Feig2011,miklos2011,Wang2012b}. Additionally, crowding affects the structure and dynamics of water around proteins, which has an indirect but important effect in their biological activities~\cite{Harada2012a,king2013}.

@@

As the available computational power increases, models become increasingly complex and are better representations of reality. The most simple class of computational models apply Brownian Dynamics calculations to probe macromolecules immersed in a solution of crowders that are treated as hard spheres with varying radii to represent different types of molecules~\cite{Ando2010}. These models usually use an implicit representation of the solvent~\cite{Mcguffee2010}. More recently, all-atom explicit solvent molecular dynamics simulations have been employed on crowded systems composed of several copies of one or a few macromolecules~\cite{Wang2017c}. However, these models doesn't reflect the true heterogeneity of biologically relevant crowded environments such as the cytoplasm. This was addressed by the development of a detailed model for the cytoplasm of the bacterium \textit{Mycoplasma genitalium} containing 103 million atoms~\cite{Feig2015}. This model, as well as a smaller version with 13 million atoms, was submitted to molecular dynamics simulations~\cite{Yu2016a}. Despite the size of these systems, the simulation times were too short, \SI{20}{\nano\second} for the complete model and \SI{140}{\nano\second} for the smaller version, to accurately characterize dynamical effects that can take place in such a large scale.

@@

Analyses were made GROMACS tools after removing artifacts generated by periodic boundary conditions from the trajectories. The Mean Square Displacement (MSD) analysis was performed to calculate the translational diffusion coefficient, which were extracted by a linear fit to MSD by averaging blocks with a length of $10 \,ns$~\cite{Allen1987a}. In principle, diffusion coefficient needs to be corrected for finite size effects~\cite{Yeh2004} but due to relatively large simulation boxes this correction is negligible. Root Mean Square Deviation (RMSD) and Root Mean Square Fluctuation (RMSF) calculations considered only C$\upalpha$ atoms and the reference frame for RMSF was the midpoint of the time interval used for analysis in each trajectory. The chain interface area of a given oligomer in a given frame was calculated as the difference between the solvent exposed surface area (SASA) of the oligomer in that frame and the sum of the SASAs of each of its chains independently in that same frame. Other analyses, such as box volume and Radial Distribution Function (RDF), followed standard procedures.

@@

\subsubsection{Structural dynamics}
Proteins are slightly more flexible in the crowded environment. Root mean square fluctuation (RMSF) calculations show that the flexibility profile of the proteins is not affected by crowding, which is in agreement with our results showing that the structural integrity of the proteins is not significantly affected in the cytoplasm model (Fig.~\ref{fig:rmsf}). On the other hand, there are some little differences between the dilute and crowding conditions. In most cases (TufA, MetE, AhpC, CspC and Ppa), proteins are slightly more flexible on the cytoplasm model (Fig.~\ref{fig:rmsf}). This is in agreement with data showing that nonspecific interactions with crowders can lead to destabilization of proteins~\cite{Feig2011, miklos2011, Wang2012b}.

@@

In this work we built an all-atom explicit solvent cytoplasm model composed by proteins, tRNA, metabolites, inorganic ions and water that sums 1.5 million atoms (Table \ref{tbl:soup_components}). The identity and amount of each component of the model was chosen according to experimental data about the composition of the cytoplasm of \textit{E. coli}~\cite{Dong1996,Bennett2009,Link1997,Mcguffee2010}.

We submitted this model to a total of \SI{3}{\micro\second} of molecular dynamics simulations using a modified AMBER FF99SB force field that was tuned to reproduce experimental data of systems composed by several macromolecules~\cite{Best2014a}. Our model was validated by reproducing experimental data about the reduction of the translational diffusion coefficient of proteins in the cytoplasm of \textit{E. coli} when compared to the dilute condition (Figure \ref{fig:translational_diffusion})~\cite{Elowitz1999,Konopka2006}. Additionally, our data is in agreement with recent simulation and experimental data that show that the unspecific interactions between macromolecules in a crowded environment can destabilize them~\cite{Feig2011,miklos2011,Wang2012b}. Specifically, we can see that, despite the overall stability of our cytoplasm model even in the \SI{}{\micro\second} timescale, the proteins are slightly less stable in  the cytoplasm model than in dilute condition (Figures \ref{fig:structural_integrity_chain} and \ref{fig:rmsf}).

>>>>

\begin{abstract}
Biomolecular crowding affects the biophysical and biochemical behavior of macromolecules when compared to the dilute environment present in experiments made with isolated proteins. Computational modeling and simulation are essential tools in the investigation of how crowding affects the structural dynamics and biological activities of macromolecules. As computational power increased, modeling and simulating large scale all-atom explicit solvent models of the prokaryote cytoplasm become possible. In this work, we build an atomistic model of the cytoplasm of \textit{Escherichia coli} composed of 1.5 million atoms and submit it to a total of \SI{3}{\micro\second} of molecular dynamics simulations, which is unprecedented for a system of this scale. Our model reproduces experimental data about the diffusion coefficient and stability of macromolecules under crowded conditions. We provide a Python script and a set of files that enables any researcher to build their own \textit{E. coli} cytoplasm models to address questions related to crowding.
%For Nature, the abstract is really an introductory paragraph set
%in bold type.  This paragraph must be ``fully referenced'' and
%less than 180 words for Letters.  This is the thing that is
%supposed to be aimed at people from other disciplines and is
%arguably the most important part to getting your paper past the
%editors.  End this paragraph with a sentence like ``Here we
%show...'' or something similar.

@@

The intracellular environment is markedly different from the \textit{in vitro} simplification from which most of our current knowledge about the biochemical and biophysical behavior of macromolecules was derived~\cite{Feig2017a}. Instead of simple and diluted, the \textit{in vivo} microenvironment is highly heterogeneous and concentrated, leading to spatial constrains that directly affects biological activities~\citep{ostrowska2019}. This effect, called macromolecular crowding, directly influences the stability of macromolecules through an equilibrium between stabilization of their native structure due to the entropic component given by the excluded volume effect~\cite{cheung2005} and destabilizationg due to enthalpic component given by unspecific interactions with the crowders~\cite{Feig2011,miklos2011,Wang2012b}. Additionally, crowding affects the structure and dynamics of water around proteins, which has an indirect but important effect in their biological activities~\cite{Harada2012a,king2013}.

@@

As the available computational power increases, models become increasingly complex and are better representations of reality. The most simple class of computational models apply Brownian Dynamics calculations to probe macromolecules immersed in a solution of crowders that are treated as hard spheres with varying radii to represent different types of molecules~\cite{Ando2010}. These models usually use an implicit representation of the solvent~\cite{Mcguffee2010}. More recently, all-atom explicit solvent molecular dynamics simulations have been employed on crowded systems composed of several copies of one or a few macromolecules~\cite{Wang2017c}. However, these models do not reflect the true heterogeneity of biologically relevant crowded environments such as the cytoplasm. This was addressed by the development of a detailed model for the cytoplasm of the bacterium \textit{Mycoplasma genitalium} containing 103 million atoms~\cite{Feig2015}. This model, as well as a smaller version with 13 million atoms, was submitted to molecular dynamics simulations~\cite{Yu2016a}. Despite the size of these systems, the simulation times were too short, \SI{20}{\nano\second} for the complete model and \SI{140}{\nano\second} for the smaller version, to accurately characterize dynamical effects that can take place in such a large scale.

@@

Analyses were done by GROMACS tools after removing artifacts generated by periodic boundary conditions from the trajectories. The Mean Square Displacement (MSD) analysis was performed to calculate the translational diffusion coefficient, which were extracted by a linear fit to MSD by averaging blocks with a length of $10 \,ns$~\cite{Allen1987a}. In principle, diffusion coefficient needs to be corrected for finite size effects~\cite{Yeh2004} but due to relatively large simulation boxes this correction is negligible. Root Mean Square Deviation (RMSD) and Root Mean Square Fluctuation (RMSF) calculations considered only C$\upalpha$ atoms and the reference frame for RMSF was the midpoint of the time interval used for analysis in each trajectory. The chain interface area of a given oligomer in a given frame was calculated as the difference between the solvent exposed surface area (SASA) of the oligomer in that frame and the sum of the SASAs of each of its chains independently in that same frame. Other analyses, such as box volume and Radial Distribution Function (RDF), followed standard procedures.

@@

\subsubsection{Structural dynamics}
Proteins are slightly more flexible in the crowded environment. Root mean square fluctuation (RMSF) calculations show that the flexibility profile of the proteins is not affected by crowding, which is in agreement with our results showing that the structural integrity of the proteins is not significantly affected in the cytoplasm model (Fig.~\ref{fig:rmsf}). On the other hand, there are some little differences between the dilute and crowding conditions. In most cases (TufA, MetE, AhpC, CspC and Ppa), proteins are slightly more flexible in the cytoplasm model (Fig.~\ref{fig:rmsf}). This is in agreement with data showing that nonspecific interactions with crowders can lead to destabilization of proteins~\cite{Feig2011, miklos2011, Wang2012b}.

@@

In this work we built an all-atom explicit solvent cytoplasm model composed of proteins, tRNA, metabolites, inorganic ions and water that sums 1.5 million atoms (Table \ref{tbl:soup_components}). The identity and amount of each component of the model was chosen according to experimental data about the composition of the cytoplasm of \textit{E. coli}~\cite{Dong1996,Bennett2009,Link1997,Mcguffee2010}.

We submitted this model to a total of \SI{3}{\micro\second} of molecular dynamics simulations using a modified AMBER FF99SB force field that was tuned to reproduce experimental data of systems composed of several macromolecules~\cite{Best2014a}. Our model was validated by reproducing experimental data about the reduction of the translational diffusion coefficient of proteins in the cytoplasm of \textit{E. coli} when compared to the dilute condition (Figure \ref{fig:translational_diffusion})~\cite{Elowitz1999,Konopka2006}. Additionally, our data is in agreement with recent simulation and experimental data that show that the unspecific interactions between macromolecules in a crowded environment can destabilize them~\cite{Feig2011,miklos2011,Wang2012b}. Specifically, we can see that, despite the overall stability of our cytoplasm model even in the \SI{}{\micro\second} timescale, the proteins are slightly less stable in  the cytoplasm model than in dilute condition (Figures \ref{fig:structural_integrity_chain} and \ref{fig:rmsf}).
